Who does big team science?

### 2024-12-05 manuscript_definition.Rmd
updated with RQ1 results

# things to import
var_list <- c("author_stats", "paper_stats", "author_paper_year")

@@

-   Research Question 1: Using current publication values, what should define big team science versus team science? 
  - Question 1A: What number of authors and institutional affiliations should designate the differences between team science and big team science?
  - Question 1B: Using the definition from 1A, are there changes in the number of publications over time?

  ### come back to this ###
-   Research Question 2: How has the diversity of those involved in big team science versus team science changed over time?

@@

We used data from 1970 and forward in the Scopus database, as it is noted online
that this time period includes cited references for calculation of
several of our variables described below. We analyzed our results
based on four subject areas present in the Scopus database: Physical
Sciences, Health Sciences, Social Sciences, and Life Sciences. We
filtered the database to include articles, articles in press, business
articles, conference papers, data papers, preprints, and surveys using
Elsevier's classification system. This project was supported by access
to the Scopus database through the International Center for the Study of
Research.

@@

# one row per author per paper with all authors with at least two  
team_author_aff_combo <- 196835398
team_author_aff_papers <- 32454393
team_author_aff_author <- 28353445
team_author_aff_subject_noNA <- 241269297
team_author_aff_papers_noNA <- 32448373
team_author_aff_author_noNA <- 28350468
```

WRITE THE THING --> mention max in the text because too crazy for the graph

```{r stats-authors}
temp_graph <- DF$author_stats %>% 
  filter(SubjectArea != "Multidisciplinary")

@@

WRITE THE THING --> make this a table to import

@@

```{r big-teams, results = 'asis'}
big_teams_papers <- 968765
big_teams_authors <- 4541369

kable(DF$paper_stats %>% 
        arrange(Statistic))
```

@@

```{r}
temp <- DF$author_paper_year %>% 
  filter(SubjectArea != "Multidisciplinary") %>% 
  filter(Year < 2024)

@@

p1 <- ggplot(temp %>% 
         filter(group == "regular"), aes(x = Year, y = Number_Papers, colour = SubjectArea)) + 
  geom_point() + 
  theme_minimal() + 
  theme(legend.position = "none")

p2 <- ggplot(temp %>% 
         filter(group == "big"), aes(x = Year, y = Number_Papers, colour = SubjectArea)) + 
  geom_point() + 
  theme_minimal() + 
  theme(legend.position = "bottom")

p1 | p2
```

```{r}
group <- unique(temp$group)
subjarea <- unique(temp$SubjectArea)
corr.results <- list()

for (group_label in group){
  for (SA in subjarea){
    stuff <- temp %>% 
      filter(group == group_label) %>% 
      filter(SubjectArea == SA) 
    # Attempt with nlsLM for more robust optimization
    fit <- nlsLM(Number_Papers ~ a * exp(r * Year), 
             start = list(a = 0.2, r = 0.05), 
             data = stuff)
    corr.results[[paste0(group_label, "_", SA)]] <- tidy(fit)

  }
}

corr.DF <- bind_rows(corr.results) %>% 
  mutate(group = rep(group, each = 8)) %>% 
  mutate(SubjectArea = c(
    rep(subjarea, each = 2),
    rep(subjarea, each = 2)
    ))

ggplot(corr.DF, aes(x = SubjectArea, y = estimate, color = group)) + 
  geom_point() +
  theme_minimal() + 
  geom_linerange(aes(ymin = conf.low, ymax = conf.high))
```

In the exponential model ￼, the parameters ￼ and ￼ have specific interpretations:
	1.	Parameter ￼ (Initial Value or Scale):
	•	￼ represents the initial value or the scaling factor of the exponential function. It indicates the starting level of the dependent variable (in this case, Number_Papers) when the Year is at its baseline (e.g., the first year in your dataset or zero if Year is centered).
	•	A larger value of ￼ would mean a higher starting point for the number of papers.
	2.	Parameter ￼ (Growth Rate):
	•	￼ is the growth rate parameter. It indicates how quickly the number of papers grows over time.
	•	If ￼ is positive, the model shows exponential growth, meaning the number of papers increases over time at an accelerating rate.
	•	If ￼ is negative, the model would represent exponential decay, suggesting that the number of papers decreases over time.
	•	The magnitude of ￼ affects the steepness of the growth or decay curve, with larger values (in absolute terms) indicating a more rapid change.

Interpretation Example:
	•	If ￼ and ￼, then:
	•	The starting number of papers is about 2 (when Year is zero or the baseline year).
	•	The number of papers grows by about ￼ per unit of Year, as ￼ implies a growth rate of ￼ per year.
	•	If ￼, the model would predict a constant number of papers over time, as the exponential term ￼ becomes 1.

Let me know if you’d like help with the interpretation for specific values in your fitted model.

```{r total-authors}
# updated
total_authors <- 3047067
avg_authors <- apa_num(49.30516877182394)
sd_authors <- apa_num(212.9786134634032)
min_authors <- 10
med_authors <- 18
max_authors <- 5568
```

The total number of unique authors across all publications was
`r total_authors`. The mean number of authors per publication was *M* =
`r avg_authors` (*SD* = `r sd_authors`, *Med* = `r med_authors`) with a
range of `r min_authors` to `r max_authors`. The median and average
number of authors by subject area are displayed in Figure
\@ref(fig:fig-author-year). In general, the average and median number of
authors increased over time, with the exception of the skew in the
Physical Sciences. Interestingly, the effect in the Physical Sciences
appears to be declining toward the general trends seen in other areas in
the last few decades.

## data split

- "teams": 2-10 authors, 2-4 inst
- "big ass": 11+, 5+

## RQ1 change over time

pattern of author numbers over time --> just the median because of the data split 
four panel figure one for each area --> with two sets of dots for teams, "team", and BAT 
since there's a restriction of range, do we need to calculate based on the proportion of the data or proportion of the range that it could increase

```{r fig-author-year, include = TRUE, fig.cap="Number of authors included on big-team science papers per year by subject area. Given the large skew in the data, the left panel presents the median number of authors per manuscript, and the right panel presents the average number of authors per manuscript by year.", fig.width=10, fig.height=6}
avg_author_time <- 
  DF$author_time %>% 
  filter(SubjectArea != "Multidisciplinary") %>% 
  # filter(Number_Papers >= 3) %>% 
  mutate(low = (Average_Num_Authors - as.numeric(SD_Num_Authors)), 
         high = (Average_Num_Authors + as.numeric(SD_Num_Authors)))

# ggplot(avg_author_time, aes(Year, Med_Num_Authors)) + 
#   theme_classic() + 
#   geom_point(aes(size = Number_Papers)) + 
#   geom_point(aes(Year, Average_Num_Authors, color = "red")) + 
#   #geom_pointrange(aes(ymin = low, ymax = high)) + 
#   facet_wrap(~SubjectArea) + 
#   ylab("Average Number of Authors") + 
#   xlab("Publication Year") + 
#   theme(legend.position = "bottom") + 
#   guides(color = "none") + 
#   scale_size_continuous(name = "Number Papers")

median_authors <- 
  ggplot(avg_author_time, aes(Year, Med_Num_Authors, color = SubjectArea)) + 
  theme_classic(base_size = 15) + 
  geom_point() + 
    ylab("Median Number of Authors") + 
    xlab("Publication Year")

mean_authors <- 
  ggplot(avg_author_time, aes(Year, Average_Num_Authors, color = SubjectArea)) + 
  theme_classic(base_size = 15) + 
  geom_point() + 
    ylab("Average Number of Authors") + 
    xlab("Publication Year")

median_authors + 
  scale_color_brewer(palette = "Spectral", name = "Subject Area") + 
  mean_authors + 
  scale_color_brewer(palette = "Spectral", name = "Subject Area") + 
  plot_layout(guides = "collect") & theme(legend.position = 'bottom')

ggsave(filename = "figure/figure_3_num_authors.png", width = 10, height = 6)
```

## RQ4: Author Diversity Statistics

@@

Figure \@ref(fig:fig-career) portrays the average career length for
authors involved in BTS publications across years. Career length was
defined as the year of first publication minus the current year, and
higher numbers mean longer careers. To analyze trends over time, we
calculated the average career length for each publication (i.e., average
author career lengths to create one score for each paper) and analyzed a
regression analysis using career length to predict year of publication.
In order to show variance between individuals, we calculated the
standard deviation of career length for each publication and used this
variance as an additional predictor.

@@

@@

```{r fig-career, include = TRUE,  warning = FALSE, message = FALSE, fig.cap = "Average career length for big-team science authors. Larger dots indicate more variability in career length for authors by averaging the standard deviation in career length for each manuscript within a year. The data has been filtered to at least 10 publications in a year for this graph.", fig.width=10, fig.height=6}
temp <- ggplot(DF$career_time %>% 
         filter(SubjectArea != "Multidisciplinary") %>% 
         filter(num_papers >= 10) %>% 
         filter(Year < 2024), 
       aes(Year, avg_career, color = SubjectArea)) +
  theme_classic(base_size = 15) + 
  geom_point(aes(size = sd_career)) +
  # geom_point() + 
  # facet_wrap(~SubjectArea) + 
  theme(legend.position = "bottom") + 
  ylab("Average Career Length") +
  xlab("Year of Publication") + 
  scale_color_brewer(palette = "Spectral", name = "Subject Area") +
  guides(size = "none")  
  #scale_size_continuous(name = "Number Papers")

suppressWarnings(suppressMessages(temp))

@@

```{=tex}
\begingroup
\setlength{\parindent}{-0.5in}

```
::: {#refs custom-style="Bibliography"}
:::

```{=tex}
\endgroup
```
\newpage

>>>>

# things to import
var_list <- c("author_stats", "paper_stats", "author_paper_year",
              "career_time")

@@

- Research Question 1: Exploring historical and current publication
    values, what should define big team science versus team science?
  - Question 1A: What number of authors and institutional affiliations
    should designate the differences between team science and big team
    science?
  - Question 1B: Using the definition from 1A, are there changes in the
    number of publications over time?
- Research Question 2: How has the diversity of those involved in big
    team science versus team science changed over time?

@@

We used data from 1970-2024 in the Scopus database, as it is noted
online that 1970 and forward includes cited references for calculation
of several of our variables described below. We analyzed our results
based on four subject areas present in the Scopus database: Physical
Sciences, Health Sciences, Social Sciences, and Life Sciences. We
filtered the database to include articles, articles in press, business
articles, conference papers, data papers, preprints, and surveys using
Elsevier's classification system. This project was supported by access
to the Scopus database through the International Center for the Study of
Research.

@@

# one row per author per paper with all authors with at least two  
team_author_aff_combo <- 196835398
team_author_aff_papers <- 32454393
team_author_aff_author <- 28353445

team_author_aff_subject_noNA <- 241269297
team_author_aff_papers_noNA <- 32448373
team_author_aff_author_noNA <- 28350468
```

The total number of papers included in the Scopus database at the time
of this analysis was `r total_papers`. `r total_papers_past1970`
articles were included past 1970 in the defined article types, which
included `r distinct_authors_past1970` distinct authors. We then
filtered the data to include only teams, which was defined as two
authors from at least two institutions. The total number of papers for
team projects was `r team_author_aff_papers` and
`r team_author_aff_author` distinct authors. The data was then
classified into subject areas by paper, which lead to missing data. The
final number of papers included was `r team_author_aff_papers_noNA` with
`r team_author_aff_author_noNA` distinct authors. The dataset was
curated to include one row per author, paper, and subject area (i.e.,
long format [@wickham2007]) which included `r team_author_aff_subject_noNA` total rows of data. For analyses and descriptive statistics below, we include sample sizes for clarity.

```{r stats-authors-fig, include = F, fig.cap="The left panel depicts the number of authors included on a paper by subject area, and the right panel demonstrates the number of affiliations by subject area. The boxplot shows the median (bold line), the interquartile range (the box), and the minimum to the 90th percentile of the number of authors/affiliations as the range line. Normally these plots include the entire range of the data, but these extreme range made the boxplot information unreadable. The dots indicate the average number of authors/affiliations for each area with the size of the dot indicating the standard deviation of the statistic. Therefore, larger dots indicate more variability in the number of authors and affiliations. "}
temp_graph <- DF$author_stats %>% 
  filter(SubjectArea != "Multidisciplinary")

@@

\@ref(fig:stats-authors-fig) displays the number of authors and affiliations by subject area. The figure demonstrates that the median number of authors is largest for health sciences, followed by life science, physical sciences, and then social sciences. The general pattern of team authorship includes about 2-8 authors, from about 2-4 institutions. We used the maximum (i.e., across all subject areas) 90th percentile as our exploratory definition for big team papers after examining the results from this analysis. We selected this percentile to have the high of the distribution, but also to be able to include enough papers for analysis across time. Therefore, big teams were defined as 11 authors from at least 6 different institutions[^In a previous version of this manuscript, we defined big teams as 10 authors from at least 10 institutions based on our own experiences working with research consortium. All definitions are likely subjective, but the definition in this manuscript represents the top 10% of author and affiliations in a large body of papers].

@@

```{r big-teams-table, results = 'asis'}
big_teams_papers <- 968765
big_teams_authors <- 4541369

apa_table(DF$paper_stats %>% 
        arrange(Statistic), 
        caption = "Number of Authors and Papers by Subject Area",
        note = "Papers can be classified into multiple categories.", 
        align = c("l", rep("c", 5)), 
        digits = 0)
```

\@ref(tab:big-teams-table) includes the number of distinct authors and papers for each subject area by overall teams and big teams using our 90th percentile definition. The total number of distinct authors for big team papers was `r big_teams_papers` with `r big_teams_authors` distinct authors. In RQ2, we split the big team data into small (6-10), medium, and large big team grouping for convenience to display/analyze geopolitical regions. The table shows the number of authors and papers for those analyses.

@@

For analyzing changes across time, we split the data into regular teams (2-10 authors, 2-5 affiliations) and big teams (as defined above, 11+ authors, 6+ affiliations). The number of papers found in Scopus across time for each subject area are displayed in Figure \@ref(paper-time-fig). The visual results indicated that the number of regular team papers was increasing the most in physical sciences for all manuscripts, followed by life and health sciences, and the last is social sciences. Examining only big teams shows that the rate is also increasing across time. All teams appear to start increasing in the 1990s, while big teams do not start increasing off floor effects until past 2000. The health and life sciences show the largest increases across time in big teams with the smallest trend in the social sciences.

```{r paper-time-fig, include = F, fig.cap = "The number of manuscripts across time for all team science papers (left) and big team science papers (right)."}
temp <- DF$author_paper_year %>% 
  filter(SubjectArea != "Multidisciplinary") %>% 
  filter(Year < 2024) %>% 
  mutate(SubjectArea = gsub("Sciences", "", SubjectArea))

@@

p1 <- ggplot(temp %>% 
         filter(group == "regular"), 
         aes(x = Year, y = Number_Papers, 
             colour = SubjectArea)) + 
  geom_point() + 
  theme_minimal() + 
  scale_color_discrete(name = "Subject Area") + 
  theme(legend.position = "bottom",
        axis.text.y = element_text(angle = 45, hjust = 1)) + # Angles the y-axis text
  ylab("Number of Papers") + 
  ggtitle("Regular Teams")

p2 <- ggplot(temp %>% 
         filter(group == "big"), 
         aes(x = Year, y = Number_Papers, 
             colour = SubjectArea)) + 
  geom_point() + 
  theme_minimal() + 
  scale_color_discrete(name = "Subject Area") + 
  theme(legend.position = "bottom",
        axis.text.y = element_text(angle = 45, hjust = 1)) + # Angles the y-axis text
  ylab(NULL) +  # Removes the y-axis label
  ggtitle("Big Teams")

combined_plot <- p1 + p2 +
  plot_layout(guides = "collect") & # Collect shared legend
  theme(legend.position = "bottom")  # Center the legend at the bottom

combined_plot
```

```{r rate-estimates-fig, include = F, fig.cap = "Exponential growth rate estimates with 95% confidence intervals."}
# Initialize the results list
corr.results <- list()

# Ensure 'group' and 'subjarea' are defined
group <- unique(temp$group)
subjarea <- unique(temp$SubjectArea)

# Loop through groups and subject areas
for (group_label in group) {
  for (SA in subjarea) {
    stuff <- temp %>%
      filter(group == group_label) %>%
      filter(SubjectArea == SA)

    if (nrow(stuff) < 3 || any(stuff$Number_Papers <= 0)) {
      message("Skipping ", group_label, " - ", SA, ": insufficient or invalid data")
      next
    }

    # Calculate starting values
    a_start <- mean(stuff$Number_Papers, na.rm = TRUE)
    log_y <- log(stuff$Number_Papers)
    b_start <- coef(lm(log_y ~ stuff$Year))[2]

    tryCatch({
      fit <- nlsLM(
        Number_Papers ~ a * exp(b * Year), 
        data = stuff, 
        start = list(a = a_start, b = b_start)
      )
      corr.results[[paste0(group_label, "_", SA)]] <- tidy(fit)
    }, error = function(e) {
      message("Fit failed for ", group_label, " - ", SA, ": ", e$message)
    })
  }
}

corr.DF <- bind_rows(corr.results) %>% 
  mutate(group = rep(group, each = 8)) %>% 
  mutate(SubjectArea = c(
    rep(subjarea, each = 2),
    rep(subjarea, each = 2)
    )) %>% 
  filter(term != "a") %>% 
  mutate(conf.low = estimate-2*std.error,
         conf.high = estimate+2*std.error)

ggplot(corr.DF, aes(x = SubjectArea, y = estimate, color = group)) + 
  geom_point() +
  theme_minimal() + 
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = .2) + 
  ylab("Rate Estimate") + 
  xlab("Subject Area") + 
  scale_color_discrete(name = "Team Size", 
                       labels = c("Big", "Regular")) + 
  theme(legend.position = "bottom") + 
  geom_hline(yintercept = 0)
```

Using the `minpack.lm` library [CITE ME], we calculated the exponential rate of growth for regular teams and big teams, and these results are shown in \@ref(fig:rate-estimates-fig). All growth rate confidence intervals excluded zero, indicating an exponential increase in the number of team papers over time. Big team growth rates were always higher than their regular team counterparts, but the 95% confidence intervals for the growth estimate overlapped for all statistics. Therefore, the growth trends, while visually appearing to be different, were likely similar for each subject area and team size when examined by estimating exponential growth statistics.

## RQ2: Diversity Statistics

@@

Figure \@ref(fig:fig-career) portrays the average career length for
authors involved in BTS publications across years. Career length was
defined as the year of first publication minus the current year, and
higher numbers mean longer careers. To analyze trends over time, we
calculated the average career length for each publication (i.e., averaging author career lengths to create one score for each paper) and analyzed a
regression analysis using career length to predict year of publication.
In order to show variance between individuals, we calculated the
standard deviation of career length for each publication and used this
variance as an additional predictor.

@@

### Edit me

@@

```{r fig-career, include = TRUE,  warning = FALSE, message = FALSE, fig.cap = "Average career length for big-team science authors. Larger dots indicate more variability in career length for authors by averaging the standard deviation in career length for each manuscript within a year. The data has been filtered to at least 10 publications in a year for this graph.", fig.width=10, fig.height=6}
ggplot(DF$career_time %>% 
         filter(SubjectArea != "Multidisciplinary") %>% 
         filter(num_papers >= 10) %>% 
         filter(Year < 2024) %>% 
         mutate(Team = factor(Team, 
                              levels = c("Regular", "Big"))), 
       aes(Year, avg_career, color = Team)) +
  theme_classic(base_size = 15) + 
  geom_point(aes(size = sd_career)) +
  # geom_point() + 
  facet_wrap(~SubjectArea) + 
  theme(legend.position = "bottom") + 
  ylab("Average Career Length") +
  xlab("Year of Publication") + 
  scale_color_brewer(palette = "Dark2", name = "Subject Area") +
  guides(size = "none")  
  #scale_size_continuous(name = "Number Papers")

@@

\begingroup
\setlength{\parindent}{-0.5in}

::: {#refs custom-style="Bibliography"}
:::

\endgroup
